---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


```{r}
activity <- read.csv("activity.csv")

activity$date <- as.Date(activity$date) # formatting date as date



acti_clear<- complete.cases(activity)

cleared_activity<- activity[acti_clear,] #clearing NA rows





```

## What is mean total number of steps taken per day?

Make a histogram of the total number of steps taken each day

```{r}

dailystep <- aggregate(cleared_activity[,1],list(date= cleared_activity$date),sum)

hist(dailystep$x, xlab="total steps per day", main= "Histogram of steps per day")

```


Calculate and report the mean and median total number of steps taken per day
```{r}

#calculating mean first
agg_tbl <-  aggregate(cleared_activity[,1],list(date= cleared_activity$date),mean)

#median next
agg_tbl_med <- aggregate(cleared_activity[,1],list(date= cleared_activity$date),median)

#combining them
agg_tbl<- cbind(agg_tbl,agg_tbl_med[,2])

#rename colums
colnames(agg_tbl)[c(2,3)] <- c("mean", "median")

agg_tbl
```


## What is the average daily activity pattern?
Make a time series plot  of the 5-minute interval and the average number of steps taken, averaged across all days
```{r}

agg_int <-  aggregate(cleared_activity[,1],list(interval= cleared_activity$interval),mean)

with(agg_int,plot(interval,x,type="l", ylab="daily average for interval"))
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

max_step<- agg_int[agg_int$x== max(agg_int[,2]),1]

max_step
```


## Imputing missing values

Calculate and report the total number of missing values in the dataset

```{r}

nrow(activity[!acti_clear,])

```
Devise a strategy for filling in all of the missing values in the dataset: Missing values will be calculated based on average interval value

```{r}


ext_activ<- merge(activity, agg_int)
ext_activ[!complete.cases(ext_activ),2]<- ext_activ[!complete.cases(ext_activ),4]
new_activity <- ext_activ[,1:3]

```

Make a histogram of the total number of steps taken each day with the missing values being replaced

```{r}

new_dailystep <- aggregate(new_activity[,2],list(date= new_activity$date),sum)

hist(new_dailystep$x, xlab="total steps per day after missing values added",
     main= "Histogram of steps per day")

```


Calculate and report the mean and median total number of steps taken per day after missing data filled in
```{r}

#calculating mean first
new_agg_tbl <-  aggregate(new_activity[,2],list(date= new_activity$date),mean)

#median next
new_agg_tbl_med <- aggregate(new_activity[,2],list(date= new_activity$date),median)

#combining them
new_agg_tbl<- cbind(new_agg_tbl,new_agg_tbl_med[,2])

#rename colums
colnames(new_agg_tbl)[c(2,3)] <- c("mean", "median")

new_agg_tbl
```


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
set_day <- function(x) {

    if (weekdays(as.Date(x)) == "Sunday" | weekdays(as.Date(x)) == "Saturday") 
       day <- "weekend"
    else
        day <- "weekday"
        
    day    
}

new_activity$wkday <- as.factor(sapply(new_agg_tbl[,1],set_day))

```


Make a panel plot containing a time series plot  of the 5-minute interval (x-axis) and the average number of steps taken,
```{r}

par(mfrow = c(2, 1))

new_agg_int_wkday <-  aggregate(steps~interval,new_activity,mean, subset= new_activity$wkday=="weekday")

with(new_agg_int_wkday,plot(interval,steps,type="l", main="weekday average for interval"))

new_agg_int_wkend <-  aggregate(steps~interval,new_activity,mean, subset= new_activity$wkday=="weekend")

with(new_agg_int_wkend,plot(interval,steps,type="l", main="weekend average for interval"))    

```

